---
title: 'Pipeline ranking on immune data with blueprint as truth'
author: 'Christoph Hafemeister'
date: '`r format(Sys.time(), "%B %d, %Y %H:%M:%S %Z")`'
output:
  html_document: 
    toc: yes
    toc_float: yes
    code_folding: hide
    highlight: pygments
    df_print: paged
---

# Setup

```{r}
source("R/setup.R")
source('R/utilities.R')

library('dplyr')
library('ComplexHeatmap')
library('ggplot2')

RES_DIR <- file.path(config$out_root, 'results', 'immune_data_summarized')
FIG_DIR <- file.path(config$out_root, 'figures')
FIG_SUPP_DIR <- file.path(config$out_root, 'figures', 'supp')
dir.create(path = FIG_SUPP_DIR, showWarnings = FALSE, recursive = TRUE)
```

# Main

## Load the summarized results

```{r}
sum_res <- readRDS(file = file.path(RES_DIR, 'performance_immune_data.Rds'))
```

Make comparison names shorter
```{r}
sum_res <- mutate(sum_res,
                 comparison = stringr::str_replace(string = comparison, pattern = 'MONOCYTES', replacement = 'Mono'),
                 comparison = stringr::str_replace(string = comparison, pattern = 'B_CELLS', replacement = 'B'),
                 comparison = stringr::str_replace(string = comparison, pattern = 'T_CELLS', replacement = 'T'))
```



## Summarize the summarized results

Rank based on MCC and FDR

Rescale MCC per experiment; average the experiments

```{r}
# squish and rescale the MCC per experiment

fr <- filter(sum_res, is.na(expr_grp)) %>%
  group_by(comparison) %>%
  mutate(MCC_rescaled = scales::rescale(scales::oob_squish(MCC))) %>%
  group_by(transformation, de_method, pipeline) %>%
  summarise(time = sum(time), MCC_rescaled = mean(MCC_rescaled), Bias = mean(Bias), .groups = 'drop')

fdr_low <- filter(sum_res, expr_grp == 'low') %>%
  group_by(transformation, de_method, pipeline) %>%
  summarise(FDR = mean(FDR), .groups = 'drop') %>%
  dplyr::rename(FDR_low = FDR)

fdr_high <- filter(sum_res, expr_grp == 'high') %>%
  group_by(transformation, de_method, pipeline) %>%
  summarise(FDR = mean(FDR), .groups = 'drop') %>%
  dplyr::rename(FDR_high = FDR)

fr <- left_join(fr, fdr_low)
fr <- left_join(fr, fdr_high)

fr <- mutate(fr, score = MCC_rescaled + (1 - FDR_low) + (1 - FDR_high)) %>%
  arrange(-score) %>%
  mutate(pos = rank(-score, ties.method = 'min'))
fr
```

Show bias (# reported positives / # actual positives)
```{r}
plot(fr$Bias)
```


## Create heatmaps

Get data into wide format
```{r}
metrics_wide <- filter(sum_res, is.na(expr_grp)) %>%
                  arrange(comparison) %>%
                  tidyr::pivot_wider(
                              id_cols = c(pipeline, transformation, de_method),
                              names_from = c(comparison, comparison), 
                              values_from = c(Precision, Sensitivity)) %>%
  select(-transformation, -de_method) 
```

Save final ranking 
```{r}
saveRDS(list(fr = fr, metrics_wide = metrics_wide), 
        file = file.path(config$out_root, 'results', 'final_ranking_immune.Rds'))
select(fr, pos, transformation, de_method, MCC_rescaled, FDR_low, FDR_high, score, time) %>%
  readr::write_csv(file = file.path(config$out_root, 'results', 'final_ranking_immune_data_experiments.csv'))
```


Define the main plotting function
```{r}
library('ComplexHeatmap')
res_heatmap <- function(fr, metrics_wide, idx = 1:nrow(fr), do_rescale = TRUE,
                        filename = NULL, fig_width = 7, fig_height = 32, 
                        show_values = TRUE, 
                        show_col_names = TRUE, anno_left_full = TRUE) {
  fr <- fr[idx, ]
  
  mat <- as.matrix(left_join(fr, metrics_wide, by = 'pipeline') %>%
                   select(starts_with(c("Prec", "Sens"))))
  rownames(mat) <- as.character(fr$pipeline)
  colnames(mat) <- stringr::str_remove(colnames(mat), '^Precision_|^Sensitivity_')
  colnames(mat) <- stringr::str_replace_all(colnames(mat), pattern = '_', replacement = ' ')
  colnames(mat) <- gsub("(^|[[:space:]])([[:alpha:]])", "\\1\\U\\2", colnames(mat), perl=TRUE)
  colnames(mat) <- stringr::str_replace_all(colnames(mat), pattern = 'Vs', replacement = 'vs')
  
  # create a scaled version for the colors
  if (do_rescale) {
    mat_rescaled <- apply(mat, 2, scales::rescale)
    heatmap_legend_param <- list(
            title = 'Column-\nrescaled\nvalue', at = c(0, 1), 
            labels = c("Worst", "Best"), border = FALSE)
  } else {
    mat_rescaled <- mat
    heatmap_legend_param <- list(title = 'Value')
  }
  
  
  # create a function print the values in each heatmap cell
  cell_fun = function(j, i, x, y, width, height, fill) {
            grid.text(sprintf("%2.0f", mat[i, j]*100), x, y, 
                      gp = gpar(fontsize = 8, 
                                col = c('gray75', 'black')[(mat_rescaled[i, j] > 0.4) + 1]))
  }
  if (!show_values) {
    cell_fun = NULL
  }
  
  ta = HeatmapAnnotation(foo = anno_block(
    gp = gpar(fill = 'gray'),
    labels = c("Precision", "Sensitivity"), 
    labels_gp = gpar(col = "black", fontsize = 10)))
  
  anno_right <- rowAnnotation(
    MCC_rescaled = anno_barplot(fr$MCC_rescaled, bar_width = 1),
    FDR_low = anno_barplot(fr$FDR_low, bar_width = 1),
    FDR_high = anno_barplot(fr$FDR_high, bar_width = 1),
    Time = anno_barplot(fr$time, bar_width = 1),
    annotation_name_rot = 35,
    annotation_name_side = 'top',
    annotation_name_offset = unit(6, "points"),
    annotation_label = list(MCC_rescaled = expression("MCC"["rescaled"]), 
                         FDR_low = expression("FDR"["low"]), 
                         FDR_high = expression("FDR"["high"]), 
                         Time = expression("Time"["seconds"])),
    gap = unit(6, "points"))
  
  if (anno_left_full) {
    anno_left <- rowAnnotation(
      Rank = anno_text(fr$pos, 
                     just = "left", show_name = TRUE, 
                     width = unit(5, 'mm'), 
                     gp = gpar(fontsize = 10)),
    Transformation = anno_text(fr$transformation, 
                               just = "left", show_name = TRUE, 
                               width = unit(18, 'mm'), 
                               gp = gpar(fontsize = 10)),
    `DE method` = anno_text(fr$de_method, 
                           just = "left", show_name = TRUE,
                           width = unit(27, 'mm'), 
                           gp = gpar(fontsize = 10)), 
    PB = grepl(pattern = '\\d_(random|cluster)', fr$de_method),
    col = list(PB = c("FALSE" = "white", "TRUE" = "black")),
    gap = unit(10, "points"),
    annotation_name_rot = c(35, 35, 35, 0),
    annotation_name_side = 'top',
    annotation_name_offset = unit(6, "points"),
    show_legend = FALSE)
  } else {
    anno_left <- rowAnnotation(
      PB = grepl(pattern = '\\d_(random|cluster)', fr$de_method),
      col = list(PB = c("FALSE" = "white", "TRUE" = "black")),
      gap = unit(10, "points"),
      annotation_name_rot = 0,
      annotation_name_align = TRUE,
      annotation_name_side = 'top',
      show_legend = FALSE)
  }
  
  
  
  ch <- Heatmap(mat_rescaled, cluster_columns = FALSE, cluster_rows = FALSE,
          show_row_names = FALSE, show_column_names = show_col_names, column_names_rot = 45,
          col = hcl.colors(n = 21, palette = 'viridis'),
          name = 'Value', column_title = NULL,
          column_split = rep(c("A", "B"), each = ncol(mat)/2),
          top_annotation = ta, right_annotation = anno_right,
          left_annotation = anno_left,
          cell_fun = cell_fun,
          heatmap_legend_param = heatmap_legend_param)
  
  if (!is.null(filename)) {
    cairo_pdf(filename = filename, width = fig_width, height = fig_height)
    show(ch)
    decorate_annotation("FDR_low", {
      grid.lines(c(0.05, 0.05), c(0, nrow(fr))+0.5, 
                 gp = gpar(lty = 2, col = "red"), 
                 default.units = "native")
    })
    decorate_annotation("FDR_high", {
      grid.lines(c(0.05, 0.05), c(0, nrow(fr))+0.5, 
                 gp = gpar(lty = 2, col = "red"), 
                 default.units = "native")
    })
    dev.off()
  }
  return(ch)
}
```

Big heatmap
```{r}
ch <- res_heatmap(fr, metrics_wide, 
                  filename = file.path(FIG_DIR, 'ranking_immune_data_big.pdf'), 
                  fig_width = 7, fig_height = 48)
```
```{r}
ch <- res_heatmap(fr, metrics_wide, 
                  filename = file.path(FIG_DIR, 'ranking_immune_data_big_squished.pdf'), 
                  fig_width = 7, fig_height = 3.5, show_values = FALSE, 
                  show_col_names = FALSE, anno_left_full = TRUE)
```

PB vs non-PB
```{r}
pb = grepl(pattern = '\\d_(random|cluster)', fr$de_method)
ch <- res_heatmap(fr, metrics_wide, idx = which(pb),
                  filename = file.path(FIG_DIR, 'ranking_immune_data_big_pb.pdf'), 
                  fig_width = 6, fig_height = 16)
ch <- res_heatmap(fr, metrics_wide, idx = which(!pb),
                  filename = file.path(FIG_DIR, 'ranking_immune_data_big_sc.pdf'), 
                  fig_width = 6, fig_height = 22)
```

Create reduced heatmaps

```{r}
fr$de_method_grp1 <- stringr::str_remove_all(string = fr$de_method, pattern = '\\d+_(cluster|random)')
fr <- mutate(fr, de_method_grp1 = stringr::str_remove_all(string = de_method_grp1, pattern = '_(ns|pc|dc|pf|10k|gm)'))
table(fr$de_method_grp1)

idx <- mutate(fr, idx = 1:nrow(fr)) %>%
  group_by(de_method_grp1) %>%
  slice_min(order_by = idx) %>%
  pull(idx) %>% sort()

ch <- res_heatmap(fr, metrics_wide, idx = idx,
                  filename = file.path(FIG_DIR, 'ranking_immune_data_filter1.pdf'), 
                  fig_width = 7, fig_height = 5)

fr <- mutate(fr, de_method_grp2 = stringr::str_remove_all(string = de_method_grp1, pattern = '_(lrt|wald|exact|qlf|trend|voom)$'))
fr <- mutate(fr, de_method_grp2 = stringr::str_remove_all(string = de_method_grp2, pattern = '_(ns|pc|dc|pf|10k|gm)'))
table(fr$de_method_grp2)

idx <- mutate(fr, idx = 1:nrow(fr)) %>%
  group_by(de_method_grp2) %>%
  slice_min(order_by = idx) %>%
  pull(idx) %>% sort()

ch <- res_heatmap(fr, metrics_wide, idx = idx,
                  filename = file.path(FIG_DIR, 'ranking_immune_data_filter2.pdf'), 
                  fig_width = 7, fig_height = 6)
```


# Appendix

Runtime: `r time_diff(SETUP_TIME)`

Session Info
```{r}
sessionInfo()
```

Future plan
```{r}
future::plan()
```

More session info
```{r}
devtools::session_info()
```
