---
title: 'Pipeline ranking on simulated data'
author: 'Christoph Hafemeister'
date: '`r format(Sys.time(), "%B %d, %Y %H:%M:%S %Z")`'
output:
  html_document: 
    toc: yes
    toc_float: yes
    code_folding: hide
    highlight: pygments
    df_print: paged
---

# Setup

```{r}
source("R/setup.R")
source('R/utilities.R')

library('dplyr')
library('ComplexHeatmap')

RES_DIR <- file.path(config$out_root, 'results', 'simulated_data_summarized')
FIG_DIR <- file.path(config$out_root, 'figures')
FIG_SUPP_DIR <- file.path(config$out_root, 'figures', 'supp')
dir.create(path = FIG_SUPP_DIR, showWarnings = FALSE, recursive = TRUE)
```

Define some constants
```{r}
experiment_for_time <- '2k_vs_2k'
experiment_for_fdr_low <- '2k_vs_2k'
experiment_for_fdr_high <- '2k_vs_2k'
experiment_order <- c('20_vs_20', '200_vs_200', '2k_vs_2k', '200_vs_2k')
fewde_experiments <- paste('fewDE', experiment_order, sep = '_')
all_experiments <- c(experiment_order, fewde_experiments)
```

# Main

## Read in all results

Create two results tables, one for the experiments with 10% DE genes, ones for the experiments with 1% DE genes (fewDE).

```{r}
rds_files <- list.files(path = RES_DIR, pattern = '^performance_.*\\.Rds$', full.names = TRUE)
res_list <- lapply(rds_files, readRDS)

res <- as_tibble(data.table::rbindlist(res_list)) %>%
  mutate(experiment = factor(experiment, levels = all_experiments)) %>%
  filter(experiment %in% experiment_order)

res_fewde <- as_tibble(data.table::rbindlist(res_list)) %>%
  mutate(experiment = factor(experiment, levels = all_experiments)) %>%
  filter(experiment %in% fewde_experiments)
```

## Rank based on MCC and FDR

Rescale MCC per experiment-replicate; average the replicates; average the experiments.

Do not use the stratified performance, but instead use the metrics from all genes together.

```{r}
# squish and rescale the MCC per experiment-replicate
fr <- filter(res, is.na(expr_grp), is.na(category)) %>%
  group_by(experiment, replicate) %>%
  mutate(MCC_rescaled = scales::rescale(scales::oob_squish(MCC))) %>%
  group_by(transformation, de_method, pipeline, experiment) %>%
  summarise(MCC_rescaled = mean(MCC_rescaled), .groups = 'drop') %>%
  group_by(transformation, de_method, pipeline) %>%
  summarise(MCC_rescaled = mean(MCC_rescaled), .groups = 'drop')
fdr_low <- filter(res, expr_grp == 'low', experiment == experiment_for_fdr_low) %>%
  group_by(transformation, de_method, pipeline) %>%
  summarise(FDR = mean(FDR), .groups = 'drop') %>%
  dplyr::rename(FDR_low = FDR)
fdr_high <- filter(res, expr_grp == 'high', experiment == experiment_for_fdr_high) %>%
  group_by(transformation, de_method, pipeline) %>%
  summarise(FDR = mean(FDR), .groups = 'drop') %>%
  dplyr::rename(FDR_high = FDR)

fr <- left_join(fr, fdr_low)
fr <- left_join(fr, fdr_high)

fr <- mutate(fr, score = MCC_rescaled + (1 - FDR_low) + (1 - FDR_high)) %>%
  arrange(-score) %>%
  mutate(pos = rank(-score, ties.method = 'min'))
fr
```

```{r}
# TMP block
filter(res, transformation == 'counts', de_method %in% c('qlrt_gm', 'qlrt_gm_3_cluster', 'qlrt_gm_3_random', 'qlrt_gm_7_cluster', 'qlrt_gm_7_random'), experiment == '200_vs_200', replicate == 1, is.na(expr_grp), is.na(category)) %>% head() %>% as.data.frame()
```


## Create figures

We want to show the time as well, add it here
```{r}
# use avg time for the 2k vs 2k runs
res_time <- filter(res, is.na(expr_grp), is.na(category), experiment == experiment_for_time) %>%
  group_by(pipeline) %>%
  summarise(time = median(time_trans + time_de), .groups = 'drop')
fr <- left_join(fr, res_time)
fr
```

We want to show Prec and Sens per pipeline per experiment
```{r}
prec_sens <- group_by(res, experiment, transformation, de_method, pipeline) %>%
  summarise(n_Prec = sum(!is.na(Precision)),
            Prec = mean(Precision, na.rm = TRUE),
            n_Sens = sum(!is.na(Sensitivity)),
            Sens = mean(Sensitivity, na.rm = TRUE), .groups = 'drop')
```

Get data into wide format
```{r}
prec_sens_wide <- arrange(prec_sens, experiment) %>%
                  tidyr::pivot_wider(
                              id_cols = c(pipeline, transformation, de_method),
                              names_from = c(experiment, experiment), 
                              values_from = c(Prec, Sens)) %>%
  select(-transformation, -de_method) 
prec_sens_wide <- left_join(select(fr, pipeline), prec_sens_wide, by = 'pipeline')
```

Save final ranking 
```{r}
saveRDS(list(fr = fr, metrics_wide = prec_sens_wide), 
        file = file.path(config$out_root, 'results', 'final_ranking_simulated.Rds'))
select(fr, pos, transformation, de_method, MCC_rescaled, FDR_low, FDR_high, score, time) %>%
  readr::write_csv(file = file.path(config$out_root, 'results', 'final_ranking_simulation_experiments.csv'))
```


Define the main plotting function
```{r}
res_heatmap <- function(fr, prec_sens_wide, idx = 1:nrow(fr), do_rescale = TRUE,
                        filename = NULL, fig_width = 7, fig_height = 32, show_values = TRUE) {
  fr <- fr[idx, ]
  prec_sens_wide <- left_join(select(fr, pipeline), prec_sens_wide, by = 'pipeline')
  
  mat <- as.matrix(left_join(fr, prec_sens_wide, by = 'pipeline') %>%
                   select(starts_with(c("Prec", "Sens"))))
  rownames(mat) <- as.character(fr$pipeline)
  colnames(mat) <- stringr::str_remove(colnames(mat), '^Prec_|^Sens_')
  colnames(mat) <- stringr::str_replace_all(colnames(mat), pattern = '_', replacement = ' ')
  
  # create a scaled version for the colors
  if (do_rescale) {
    mat_rescaled <- apply(mat, 2, scales::rescale)
    heatmap_legend_param <- list(
            title = 'Column-\nrescaled\nvalue', at = c(0, 1), 
            labels = c("Worst", "Best"), border = FALSE)
  } else {
    mat_rescaled <- mat
    heatmap_legend_param <- list(title = 'Value')
  }
  
  
  # create a function to print the values in each heatmap cell
  cell_fun = function(j, i, x, y, width, height, fill) {
            grid.text(sprintf("%2.0f", mat[i, j]*100), x, y, 
                      gp = gpar(fontsize = 8, 
                                col = c('gray75', 'black')[(mat_rescaled[i, j] > 0.4) + 1]))
  }
  if (!show_values) {
    cell_fun = NULL
  }
  
  ta = HeatmapAnnotation(foo = anno_block(
    gp = gpar(fill = 'gray'),
    labels = c("Precision", "Sensitivity"), 
    labels_gp = gpar(col = "black", fontsize = 10)))
  
  anno_right <- rowAnnotation(
    MCC_rescaled = anno_barplot(fr$MCC_rescaled, bar_width = 1),
    FDR_low = anno_barplot(fr$FDR_low, bar_width = 1),
    FDR_high = anno_barplot(fr$FDR_high, bar_width = 1),
    Time = anno_barplot(fr$time, bar_width = 1),
    annotation_name_rot = 35,
    annotation_name_side = 'top',
    annotation_name_offset = unit(6, "points"),
    annotation_label = list(MCC_rescaled = expression("MCC"["rescaled"]), 
                         FDR_low = expression("FDR"["low"]), 
                         FDR_high = expression("FDR"["high"]), 
                         Time = expression("Time"["seconds"])),
    gap = unit(6, "points"))
  
  anno_left <- rowAnnotation(
    Rank = anno_text(fr$pos, 
                     just = "left", show_name = TRUE, 
                     width = unit(5, 'mm'), 
                     gp = gpar(fontsize = 10)),
    Transformation = anno_text(fr$transformation, 
                               just = "left", show_name = TRUE, 
                               width = unit(18, 'mm'), 
                               gp = gpar(fontsize = 10)),
    `DE method` = anno_text(fr$de_method, 
                           just = "left", show_name = TRUE,
                           width = unit(27, 'mm'), 
                           gp = gpar(fontsize = 10)), 
    PB = grepl(pattern = '\\dpr', fr$de_method),
    col = list(PB = c("FALSE" = "white", "TRUE" = "black")),
    gap = unit(10, "points"),
    annotation_name_rot = c(35, 35, 35, 0),
    annotation_name_side = 'top',
    annotation_name_offset = unit(6, "points"),
    show_legend = FALSE)
  
  
  ch <- Heatmap(mat_rescaled, cluster_columns = FALSE, cluster_rows = FALSE,
          show_row_names = FALSE, show_column_names = TRUE, column_names_rot = 45,
          col = hcl.colors(n = 21, palette = 'viridis'),
          name = 'Value', column_title = 'Overall performance',
          column_split = rep(c("A", "B"), each = 4),
          top_annotation = ta, right_annotation = anno_right,
          left_annotation = anno_left,
          cell_fun = cell_fun,
          heatmap_legend_param = heatmap_legend_param)
  
  if (!is.null(filename)) {
    cairo_pdf(filename = filename, width = fig_width, height = fig_height)
    show(ch)
    decorate_annotation("FDR_low", {
      grid.lines(c(0.05, 0.05), c(0, nrow(fr))+0.5, 
                 gp = gpar(lty = 2, col = "red"), 
                 default.units = "native")
    })
    decorate_annotation("FDR_high", {
      grid.lines(c(0.05, 0.05), c(0, nrow(fr))+0.5, 
                 gp = gpar(lty = 2, col = "red"), 
                 default.units = "native")
    })
    dev.off()
  }
  return(ch)
}
```


Big heatmap
```{r}
ch <- res_heatmap(fr, prec_sens_wide, 
                  filename = file.path(FIG_DIR, 'ranking_sim_data_big.pdf'), 
                  fig_width = 7, fig_height = 48)
```

Same, but without cell values and squished together
```{r}
ch <- res_heatmap(fr, prec_sens_wide, 
                  filename = file.path(FIG_DIR, 'ranking_sim_data_big_squished.pdf'), 
                  fig_width = 7, fig_height = 5, show_values = FALSE)
```


PB vs non-PB
```{r}
pb = grepl(pattern = '\\dpr', fr$de_method)
ch <- res_heatmap(fr, prec_sens_wide, idx = which(pb),
                  filename = file.path(FIG_DIR, 'ranking_sim_data_pb.pdf'), 
                  fig_width = 7, fig_height = 22)
ch <- res_heatmap(fr, prec_sens_wide, idx = which(!pb),
                  filename = file.path(FIG_DIR, 'ranking_sim_data_sc.pdf'), 
                  fig_width = 7, fig_height = 32)
```

Create reduced heatmaps


```{r}
# only the ones with low FDR_high
ch <- res_heatmap(fr, prec_sens_wide, idx = which(fr$FDR_high < 0.1),
                  filename = file.path(FIG_DIR, 'ranking_sim_data_low_FDR_high.pdf'), 
                  fig_width = 7, fig_height = 9)

```


```{r}
fr$de_method_grp1 = stringr::str_remove_all(string = fr$de_method, pattern = '\\d+(?=pr)')
fr <- mutate(fr, de_method_grp1 = stringr::str_remove_all(string = de_method_grp1, pattern = '_(ns|pc|dc|pf|10k|gm)'))
table(fr$de_method_grp1)

idx <- mutate(fr, idx = 1:nrow(fr)) %>%
  group_by(de_method_grp1) %>%
  slice_min(order_by = idx) %>%
  pull(idx) %>% sort()

ch <- res_heatmap(fr, prec_sens_wide, idx = idx,
                  filename = file.path(FIG_DIR, 'ranking_sim_data_filter1.pdf'), 
                  fig_width = 7, fig_height = 5)

fr <- mutate(fr, de_method_grp2 = stringr::str_remove_all(string = de_method_grp1, pattern = '_(lrt|wald|exact|qlf|trend|voom)$'))
fr <- mutate(fr, de_method_grp2 = stringr::str_remove_all(string = de_method_grp2, pattern = '_(ns|pc|dc|pf|10k|gm)'))
table(fr$de_method_grp2)

idx <- mutate(fr, idx = 1:nrow(fr)) %>%
  group_by(de_method_grp2) %>%
  slice_min(order_by = idx) %>%
  pull(idx) %>% sort()

ch <- res_heatmap(fr, prec_sens_wide, idx = idx,
                  filename = file.path(FIG_DIR, 'ranking_sim_data_filter2.pdf'), 
                  fig_width = 7, fig_height = 4)
```

Save final ranking 
```{r}
saveRDS(list(fr = fr, prec_sens_wide = prec_sens_wide), 
        file = file.path(RES_DIR, 'final_ranking_sim_data.Rds'))
```

One plot per transformation
```{r}
for (transf in unique(fr$transformation)) {
  idx <- which(fr$transformation == transf)
  ch <- res_heatmap(fr, prec_sens_wide, idx = idx, do_rescale = FALSE,
                  filename = file.path(FIG_SUPP_DIR, sprintf('ranking_sim_data_transf_%s.pdf', transf)), 
                  fig_width = 7, fig_height = 2 + 0.2 * length(idx))
}
```

One plot per method group
```{r}
method_grp = stringr::str_remove_all(string = fr$de_method, pattern = '\\d+(?=pr)')
method_grp = stringr::str_remove_all(string = method_grp, pattern = '_(ns|pc|dc|pf|10k|gm)')
method_grp = stringr::str_remove_all(string = method_grp, pattern = '_(emp|app)')
method_grp = stringr::str_remove_all(string = method_grp, pattern = '_(limma|presto)')
method_grp = stringr::str_remove_all(string = method_grp, pattern = '_(lrt|wald|exact|qlf)')
method_grp = stringr::str_remove_all(string = method_grp, pattern = '_(trend|voom)')
method_grp = stringr::str_remove_all(string = method_grp, pattern = 'detrate$')
table(method_grp)

for (meth in unique(method_grp)) {
  idx <- which(method_grp == meth)
  ch <- res_heatmap(fr, prec_sens_wide, idx = idx, do_rescale = FALSE,
                  filename = file.path(FIG_SUPP_DIR, sprintf('ranking_sim_data_meth_%s.pdf', meth)), 
                  fig_width = 7, fig_height = 2 + 0.2 * length(idx))
}
```


# Appendix

Runtime: `r time_diff(SETUP_TIME)`

Session Info
```{r}
sessionInfo()
```

Future plan
```{r}
future::plan()
```

More session info
```{r}
devtools::session_info()
```
