---
title: 'Run pipelines on immune data'
author: 'Christoph Hafemeister'
date: '`r format(Sys.time(), "%B %d, %Y %H:%M:%S %Z")`'
output:
  html_document: 
    toc: yes
    toc_float: yes
    code_folding: hide
    highlight: pygments
    df_print: paged
---

# Setup

```{r}
source("R/setup.R")
source('R/utilities.R')
source('R/transformations.R')
source('R/de_methods.R')
library('dplyr')
library('future.apply')
library('Matrix')

#library('Seurat')
#library('dplyr')
#library('Matrix')
#library('SingleCellExperiment')

#OUT_DIR <- file.path(config$out_root, 'simulated_data')
#dir.create(path = OUT_DIR, showWarnings = FALSE, recursive = TRUE)

DATA_DIR <- file.path(config$out_root, 'immune_data')
OUT_DIR <- file.path(config$out_root, 'results', 'immune_data')
dir.create(path = OUT_DIR, showWarnings = FALSE, recursive = TRUE)
SEED_TRANS <- 783838
SEED_DE <- 184920
```

# Main

Read the pipeline definition files
```{r}
transformations <- readr::read_csv(file = file.path(config$project_root, 'pipelines', 'transformations.csv'))
de_tests <- readr::read_csv(file = file.path(config$project_root, 'pipelines', 'de_tests.csv'))
```

Expand all the transformations
```{r}
transformations <- tidyr::separate_rows(transformations, `size factor methods`) %>% 
  mutate(idx = 1:n()) %>%
  group_by(idx) %>%
  mutate(name = stringr::str_interp(string = name, env = list(sf = `size factor methods`))) %>%
  ungroup()
```

Expand all the test methods
```{r}
de_tests <- tidyr::separate_rows(de_tests, `size factor methods`) %>% 
  tidyr::separate_rows(`pseudo replicates`) %>%
  mutate(idx = 1:n()) %>%
  group_by(idx) %>%
  mutate(name = stringr::str_interp(string = name, env = list(sf = `size factor methods`, pr = `pseudo replicates`))) %>%
  ungroup()
```

Create data frame of pipelines
```{r}
pipelines <- full_join(transformations, de_tests, by = character(), suffix = c(".trans", ".de")) %>%
  filter(`use with transformation` == 'ANY' | 
         (`use with transformation` == 'COUNTS' & grepl(pattern = 'counts', x = name.trans)) | 
         (`use with transformation` == 'SPARSE' & `result is sparse`))
```

Load the data and summarize again
```{r}
meta_data <- readRDS(file = list.files(DATA_DIR, pattern = 'metadata\\.Rds$', full.names = TRUE))
counts <- readRDS(file = list.files(DATA_DIR, pattern = 'counts\\.Rds$', full.names = TRUE))
table(meta_data$broad_cell_type)
```


We want to run the pipelines on all pairwise comparisons
```{r}
comps <- c('T CELLS vs MONOCYTES', 'T CELLS vs B CELLS', 'MONOCYTES vs B CELLS')
```

Set up the data lists that will be used by the pipelines
```{r}
comp_data <- list()
for (comp in comps) {
  group_names <- stringr::str_split(string = comp, pattern = ' vs ')[[1]]
  
  selA <- meta_data$broad_cell_type == group_names[1]
  selB <- meta_data$broad_cell_type == group_names[2]
  
  group_id <- factor(c(rep(group_names[1], sum(selA)), rep(group_names[2], sum(selB))),
                     levels = group_names)
  
  comp_counts <- cbind(counts[, selA], counts[, selB])
  
  # keep only features with at least three non-zero observations
  keep <- rowSums(comp_counts > 0) >= 3
  comp_data[[comp]] <- list(counts = comp_counts[keep, ],
                            cell_metadata = data.frame(group_id = group_id))
  
  message('comp ', comp, ': ', sum(selA), ' vs ', sum(selB), ' cells, ', sum(keep), ' features')
}
saveRDS(object = comp_data, file = file.path(DATA_DIR, 'immune_comp_data.Rds'))
```

Create list of jobs and run them
```{r}
jobs <- full_join(data.frame(comp = comps), pipelines, by = character())
message('number of jobs: ', nrow(jobs))

return_values <- future_sapply(X = 1:nrow(jobs), FUN = function(i) {
  comp <- jobs$comp[i]
  comp_for_file <- gsub(pattern = ' ', replacement = '_', x = comp)
  transformation <- jobs$name.trans[i]
  de_method <- jobs$name.de[i]
  out_file <- file.path(OUT_DIR, sprintf('%s-%s-%s.Rds', comp_for_file, transformation, de_method))
  if (file.exists(out_file)) {
    message(out_file, ' exists')
    return('exists')
  } else {
    res <- run_pipeline(input_data = comp_data[[comp]], 
                        transformation = transformation, 
                        de_method = de_method,
                        seed_trans = SEED_TRANS, seed_de = SEED_DE)
    saveRDS(object = res, file = out_file)
    return('done')
  }
}, future.seed=TRUE)
table(return_values)
```

# Appendix

Runtime: `r time_diff(SETUP_TIME)`

Session Info
```{r}
sessionInfo()
```

Future plan
```{r}
future::plan()
```

More session info
```{r}
devtools::session_info()
```
