---
title: 'Simulate data'
author: 'Christoph Hafemeister'
date: '`r format(Sys.time(), "%B %d, %Y %H:%M:%S %Z")`'
output:
  html_document: 
    toc: yes
    toc_float: yes
    code_folding: hide
    highlight: pygments
    df_print: paged
---

# Setup

```{r}
source("R/setup.R")
source('R/utilities.R')

library('Seurat')
library('dplyr')
library('Matrix')
library('SingleCellExperiment')

OUT_DIR <- file.path(config$out_root, 'simulated_data')
dir.create(path = OUT_DIR, showWarnings = FALSE, recursive = TRUE)

IMMUNE_COUNTS <- file.path(config$out_root, 'immune_data', '10k_PBMC_3p_nextgem_Chromium_Controller_filtered_feature_bc_matrix_seurat_counts.Rds')
IMMUNE_METADATA <- file.path(config$out_root, 'immune_data', '10k_PBMC_3p_nextgem_Chromium_Controller_filtered_feature_bc_matrix_seurat_metadata.Rds')
```

# Load some data to use as source for background parameters

Load counts and meta data for the immune data
```{r}
counts <- readRDS(IMMUNE_COUNTS)
md <- readRDS(IMMUNE_METADATA)
```

Subset to monocytes and create a downsampled version; make Seurat objects
```{r}
s1 <- CreateSeuratObject(counts = counts[, md$broad_cell_type == 'MONOCYTES'], project = 'mono')
s2 <- CreateSeuratObject(counts = downsample_counts(counts[, md$broad_cell_type == 'MONOCYTES'], frac = 0.85), project = 'mono_down')
```

Add cluster labels
```{r}
my_clustering <- cluster_counts(s1$RNA@counts, k = 5)
s1$cluster_id <- paste0('cluster', my_clustering)
s2$cluster_id <- paste0('cluster', my_clustering)
```


Merge the two versions
```{r}
s3 <- merge(s1, s2)
```


# Generate data using muscat

Convert Seurat object to single cell experiment
```{r}
# s3$sample_id <- s3$orig.ident
# table(s3$sample_id, s3$cluster_id)
# sce <- as.SingleCellExperiment(s3)
s3$sample_id <- paste(s3$orig.ident, s3$cluster_id, sep = '_')
s3$cluster_id <- 'cluster1'
table(s3$sample_id, s3$cluster_id)
sce <- as.SingleCellExperiment(s3)
```
Create reference from data
```{r}
ref <- muscat::prepSim(x = sce, verbose = TRUE)
head(SummarizedExperiment::rowData(ref))
```   


TMP - sim stuff
```{r}
# create one muscat ref per cluster
sce$orig_cluster <- stringr::str_extract(sce$sample_id, '\\d+$')
ref_list <- lapply(sort(unique(sce$orig_cluster)), function(cluster_name) {
  muscat::prepSim(x = sce[, sce$orig_cluster == cluster_name], verbose = TRUE)
})

# simulate per cluster
nc_group <- 2000
sim_list <- lapply(ref_list, function(ref) {
  done <- FALSE
  while (!done) {
    this_sim <- muscat::simData(ref, p_dd = c(0.8, 0, 0.2, 0, 0, 0), nc = nc_group*4, paired = FALSE)
    sid1 <- metadata(this_sim)$ref_sids[1,1] %>% stringr::str_remove_all('_cluster\\d+$')
    sid2 <- metadata(this_sim)$ref_sids[1,2] %>% stringr::str_remove_all('_cluster\\d+$')
    # we accept only specific sample orders
    if (sid1 == levels(s1$orig.ident) && sid2 == levels(s2$orig.ident)) {
      done <- TRUE
    }
  }
  print(metadata(this_sim)$ref_sids)
  return(this_sim)
})

# extract cells/genes for final data set
# for every gene, randomly choose one simulation (cluster)
gene_info <- lapply(1:length(sim_list), function(i) {
  metadata(sim_list[[i]])$gene_info %>% 
    mutate(source_sim = i)
}) %>% bind_rows()
gene_source <- group_by(gene_info, sim_gene) %>%
  summarise(source_sim = case_when(n() == 1 ~ source_sim[1], TRUE ~ sample(source_sim, size = 1))) %>%
  left_join(gene_info)
table(gene_source$source_sim, gene_source$category)
# create final count matrix
count_list <- lapply(unique(gene_source$source_sim), function(i) {
  # take the first nc_group cells per group
  sel_cells <- c(which(sim_list[[i]]$group_id == 'A')[1:nc_group], which(sim_list[[i]]$group_id == 'B')[1:nc_group])
  
  counts <- assay(sim_list[[i]], 'counts')[filter(gene_source, source_sim == i)$gene, sel_cells]
  rownames(counts) <- paste(i, rownames(counts), sep = '_')
  colnames(counts) <- paste('cell', 1:ncol(counts))
  return(counts)
})
count_mat <- do.call(rbind, count_list)
sim_gene_md <- left_join(data.frame(gene = rownames(count_mat)),  mutate(gene_source, gene = paste(source_sim, gene, sep = '_')))
sim_cell_md <- data.frame(group_id = factor(c(rep('A', times = nc_group), rep('B', times = nc_group))))

dp_out <- default_processing(counts = Matrix(count_mat), run_umap = TRUE)
md <- as.data.frame(cbind(sim_cell_md, dp_out$umap)) %>%
    dplyr::rename(UMAP_1 = '1', UMAP_2 = '2') 
ggplot(md, aes(UMAP_1, UMAP_2, group = group_id)) +
  #geom_point(aes(color = group_id, shape = stringr::str_remove(sample_id, '\\.[AB]$'))) +
  geom_point(aes(color = group_id)) +
  stat_density_2d(contour_var = 'ndensity', color = 'black', 
                  breaks = c(0.5, 0.6, 0.7, 0.8, 0.9), linewidth = 0.4, alpha = 0.5)
  
# process just one group
sel_cells <- sim_cell_md$group_id == 'A'
dp_out <- default_processing(counts = Matrix(count_mat[, sel_cells]), run_umap = TRUE)
md <- as.data.frame(cbind(sim_cell_md[sel_cells, , drop = FALSE], dp_out$umap)) %>%
    dplyr::rename(UMAP_1 = '1', UMAP_2 = '2') 
ggplot(md, aes(UMAP_1, UMAP_2, group = group_id)) +
  #geom_point(aes(color = group_id, shape = stringr::str_remove(sample_id, '\\.[AB]$'))) +
  geom_point(aes(color = group_id)) +
  stat_density_2d(contour_var = 'ndensity', color = 'black', 
                  breaks = c(0.5, 0.6, 0.7, 0.8, 0.9), linewidth = 0.4, alpha = 0.5)
# so we don't see the heterogeneity that we expect - no 5 sub-groups within the group :(
```

Second try - use the different cells
```{r}
# filter to genes that are DE in only one cluster and in all simulation sets
filtered_sim_genes <- group_by(gene_info, sim_gene) %>%
  summarise(n = n(), d_count = sum(category != 'ee'), .groups = 'drop') %>%
  filter(n == 5, d_count <= 1) %>%
  pull(sim_gene)

# sample cells

# how many cells per ref per group
ssize <- as.numeric(table(ceiling((1:nc_group) %% length(ref_list))))

count_list <- lapply(1:length(ref_list), function(i) {
  # which cells
  cellsA <- sample(which(sim_list[[i]]$group_id == 'A'), size = ssize[i])
  cellsB <- sample(which(sim_list[[i]]$group_id == 'B'), size = ssize[i])
  counts <- assay(sim_list[[i]], 'counts')[, c(cellsA, cellsB)]
  colnames(counts) <- sprintf('cell_%d_%d', i, 1:ncol(counts))
  rownames(counts) <-  metadata(sim_list[[i]])$gene_info$sim_gene
  return(counts[filtered_sim_genes, ])
})
count_mat <- do.call(cbind, count_list)

sim_groups <- as.vector(sapply(1:length(ref_list), function(i) rep(c('A', 'B'), each = ssize[i])))
sim_ref <- as.vector(sapply(1:length(ref_list), function(i) rep(i, times = ssize[i]*2)))
sim_cell_md <- data.frame(group_id = sim_groups, ref_id = factor(sim_ref))

dp_out <- default_processing(counts = Matrix(count_mat), run_umap = TRUE)
md <- as.data.frame(cbind(sim_cell_md, dp_out$umap)) %>%
    dplyr::rename(UMAP_1 = '1', UMAP_2 = '2') 
ggplot(md, aes(UMAP_1, UMAP_2, group = group_id)) +
  #geom_point(aes(color = group_id, shape = stringr::str_remove(sample_id, '\\.[AB]$'))) +
  geom_point(aes(color = group_id, shape = ref_id)) +
  stat_density_2d(contour_var = 'ndensity', color = 'black', 
                  breaks = c(0.5, 0.6, 0.7, 0.8, 0.9), linewidth = 0.4, alpha = 0.5)
  
# show how the groups are related
pb_labels <- factor(paste(md$ref_id, md$group_id))
pb <- pseudobulk(counts = count_mat, grouping = pb_labels)
cmat <- cor(log1p(pb))
ComplexHeatmap::Heatmap(cmat)
```


END TMP


Simulate data with different parameters

```{r}
p_diff <- 0.1
p_de <- p_diff
p_ep <- 0
p_dp <- 0 #p_diff / 4
p_dm <- 0 #p_diff / 2
p_db <- 0
p_dd <- c(1-p_diff, p_ep, p_de, p_dp, p_dm, p_db)
nc <- 400

N <- 13
res <- list()
for (i in 1:N) {
  this_sim <- muscat::simData(ref, p_dd = p_dd, nc = nc, paired = FALSE)
  print(table(this_sim$sample_id))
  print(metadata(this_sim)$ref_sids)
  
  # tmp
  dp_out <- default_processing(counts = Matrix(assay(this_sim, 'counts')), run_umap = TRUE)
  md <- as.data.frame(cbind(colData(this_sim), dp_out$umap)) %>%
      dplyr::rename(UMAP_1 = V1, UMAP_2 = V2) 
  ggplot(md, aes(UMAP_1, UMAP_2, group = group_id)) +
    geom_point(aes(color = group_id, shape = stringr::str_remove(sample_id, '\\.[AB]$'))) +
    #geom_point(aes(color = group_id)) +
    stat_density_2d(contour_var = 'ndensity', color = 'black', 
                    breaks = c(0.5, 0.6, 0.7, 0.8, 0.9), linewidth = 0.4, alpha = 0.5)
  gi <- metadata(this_sim)$gene_info
  head(gi)
  # end tmp
  
  res[[i]] <- list(counts = Matrix(assay(this_sim, 'counts')),
                   cell_metadata = colData(this_sim),
                   sim_metadata = metadata(this_sim)[c('experiment_info', 'n_cells', 'gene_info', 'ref_sids')])
}
saveRDS(object = res, file = file.path(OUT_DIR, 'sim_data_200_vs_200.Rds'))
```

Here we save only 10% of both groups.
```{r}
N <- 13
res <- list()
for (i in 1:N) {
  this_sim <- muscat::simData(ref, p_dd = p_dd, nc = round(nc * 0.1), ns = 1, paired = FALSE)
  print(table(this_sim$sample_id))
  print(metadata(this_sim)$ref_sids)
  
  res[[i]] <- list(counts = Matrix(assay(this_sim, 'counts')),
                   cell_metadata = colData(this_sim),
                   sim_metadata = metadata(this_sim)[c('experiment_info', 'n_cells', 'gene_info', 'ref_sids')])
}
saveRDS(object = res, file = file.path(OUT_DIR, 'sim_data_20_vs_20.Rds'))
```

Here we create ten times as many cells for both groups.
```{r}
N <- 13
res <- list()
for (i in 1:N) {
  this_sim <- muscat::simData(ref, p_dd = p_dd, nc = nc * 10, ns = 1, paired = FALSE)
  print(table(this_sim$sample_id))
  print(metadata(this_sim)$ref_sids)
  
  res[[i]] <- list(counts = Matrix(assay(this_sim, 'counts')),
                   cell_metadata = colData(this_sim),
                   sim_metadata = metadata(this_sim)[c('experiment_info', 'n_cells', 'gene_info', 'ref_sids')])
}
saveRDS(object = res, file = file.path(OUT_DIR, 'sim_data_2k_vs_2k.Rds'))
```

An unbalanced version (200 vs 2000)

```{r}
N <- 13
res <- list()
for (i in 1:N) {
  this_sim <- muscat::simData(ref, p_dd = p_dd, nc = nc * 10, ns = 1, paired = FALSE)
  sample_tab <- table(this_sim$sample_id)
  print(sample_tab)
  print(metadata(this_sim)$ref_sids)
  
  keep <- sort(sample(x = which(this_sim$sample_id == names(sample_tab)[1]), size = round(sample_tab[1] * 0.1)))
  keep <- c(keep, which(this_sim$sample_id == names(sample_tab)[2]))
  print(table(this_sim$sample_id[keep]))
  res[[i]] <- list(counts = Matrix(assay(this_sim, 'counts'))[, keep],
                   cell_metadata = colData(this_sim)[keep, ],
                   sim_metadata = metadata(this_sim)[c('experiment_info', 'n_cells', 'gene_info', 'ref_sids')])
}
saveRDS(object = res, file = file.path(OUT_DIR, 'sim_data_200_vs_2k.Rds'))
```

Repeat all simulations with fewer DE genes

```{r}
p_diff <- 0.01
p_de <- p_diff / 2
p_dp <- 0 #p_diff / 4
p_dm <- p_diff / 2
p_dd <- c(1-p_diff, 0, p_de, p_dp, p_dm, 0)
nc <- 400

N <- 13
res <- list()
for (i in 1:N) {
  this_sim <- muscat::simData(ref, p_dd = p_dd, nc = nc, ns = 1, paired = FALSE)
  print(table(this_sim$sample_id))
  print(metadata(this_sim)$ref_sids)
  
  res[[i]] <- list(counts = Matrix(assay(this_sim, 'counts')),
                   cell_metadata = colData(this_sim),
                   sim_metadata = metadata(this_sim)[c('experiment_info', 'n_cells', 'gene_info', 'ref_sids')])
}
saveRDS(object = res, file = file.path(OUT_DIR, 'sim_data_fewDE_200_vs_200.Rds'))
```

Here we save only 10% of both groups.
```{r}
N <- 13
res <- list()
for (i in 1:N) {
  this_sim <- muscat::simData(ref, p_dd = p_dd, nc = round(nc * 0.1), ns = 1, paired = FALSE)
  print(table(this_sim$sample_id))
  print(metadata(this_sim)$ref_sids)
  
  res[[i]] <- list(counts = Matrix(assay(this_sim, 'counts')),
                   cell_metadata = colData(this_sim),
                   sim_metadata = metadata(this_sim)[c('experiment_info', 'n_cells', 'gene_info', 'ref_sids')])
}
saveRDS(object = res, file = file.path(OUT_DIR, 'sim_data_fewDE_20_vs_20.Rds'))
```

Here we create ten times as many cells for both groups.
```{r}
N <- 13
res <- list()
for (i in 1:N) {
  this_sim <- muscat::simData(ref, p_dd = p_dd, nc = nc * 10, ns = 1, paired = FALSE)
  print(table(this_sim$sample_id))
  print(metadata(this_sim)$ref_sids)
  
  res[[i]] <- list(counts = Matrix(assay(this_sim, 'counts')),
                   cell_metadata = colData(this_sim),
                   sim_metadata = metadata(this_sim)[c('experiment_info', 'n_cells', 'gene_info', 'ref_sids')])
}
saveRDS(object = res, file = file.path(OUT_DIR, 'sim_data_fewDE_2k_vs_2k.Rds'))
```

An unbalanced version (200 vs 2000)

```{r}
N <- 13
res <- list()
for (i in 1:N) {
  this_sim <- muscat::simData(ref, p_dd = p_dd, nc = nc * 10, ns = 1, paired = FALSE)
  sample_tab <- table(this_sim$sample_id)
  print(sample_tab)
  print(metadata(this_sim)$ref_sids)
  
  keep <- sort(sample(x = which(this_sim$sample_id == names(sample_tab)[1]), size = round(sample_tab[1] * 0.1)))
  keep <- c(keep, which(this_sim$sample_id == names(sample_tab)[2]))
  print(table(this_sim$sample_id[keep]))
  res[[i]] <- list(counts = Matrix(assay(this_sim, 'counts'))[, keep],
                   cell_metadata = colData(this_sim)[keep, ],
                   sim_metadata = metadata(this_sim)[c('experiment_info', 'n_cells', 'gene_info', 'ref_sids')])
}
saveRDS(object = res, file = file.path(OUT_DIR, 'sim_data_fewDE_200_vs_2k.Rds'))
```



# Appendix

Runtime: `r time_diff(SETUP_TIME)`

Session Info
```{r}
sessionInfo()
```
