---
title: 'Simulate data'
author: 'Christoph Hafemeister'
date: '`r format(Sys.time(), "%B %d, %Y %H:%M:%S %Z")`'
output:
  html_document: 
    toc: yes
    toc_float: yes
    code_folding: hide
    highlight: pygments
    df_print: paged
---

# Setup

```{r}
source("R/setup.R")
source('R/utilities.R')

library('Seurat')
library('dplyr')
library('Matrix')
library('SingleCellExperiment')

OUT_DIR <- file.path(config$out_root, 'simulated_data')
dir.create(path = OUT_DIR, showWarnings = FALSE, recursive = TRUE)

IMMUNE_COUNTS <- file.path(config$out_root, 'immune_data', '10k_PBMC_3p_nextgem_Chromium_Controller_filtered_feature_bc_matrix_seurat_counts.Rds')
IMMUNE_METADATA <- file.path(config$out_root, 'immune_data', '10k_PBMC_3p_nextgem_Chromium_Controller_filtered_feature_bc_matrix_seurat_metadata.Rds')
```

# Load some data to use as source for background parameters

Load counts and meta data for the immune data
```{r}
counts <- readRDS(IMMUNE_COUNTS)
md <- readRDS(IMMUNE_METADATA)
```

Subset to monocytes and create a downsampled version; make Seurat objects
```{r}
s1 <- CreateSeuratObject(counts = counts[, md$broad_cell_type == 'MONOCYTES'], project = 'mono')
s2 <- CreateSeuratObject(counts = downsample_counts(counts[, md$broad_cell_type == 'MONOCYTES'], frac = 0.85), project = 'mono_down')
```

Merge the two versions
```{r}
s3 <- merge(s1, s2)
```


# Generate data using muscat

Convert Seurat object to single cell experiment
```{r}
s3$cluster_id <- '0'
s3$sample_id <- s3$orig.ident
sce <- as.SingleCellExperiment(s3)
```

Create reference from data
```{r}
ref <- muscat::prepSim(x = sce, verbose = TRUE)
head(SummarizedExperiment::rowData(ref))
```   

Simulate data with different parameters

```{r}
p_de <- 0.1
p_dd <- c(1-p_de, 0, p_de, 0, 0, 0)
nc <- 400

N <- 13
res <- list()
for (i in 1:N) {
  this_sim <- muscat::simData(ref, p_dd = p_dd, nc = nc, ns = 1, paired = FALSE)
  print(table(this_sim$sample_id))
  print(metadata(this_sim)$ref_sids)
  
  res[[i]] <- list(counts = Matrix(assay(this_sim, 'counts')),
                   cell_metadata = colData(this_sim),
                   sim_metadata = metadata(this_sim)[c('experiment_info', 'n_cells', 'gene_info', 'ref_sids')])
}
saveRDS(object = res, file = file.path(OUT_DIR, 'sim_data_200_vs_200.Rds'))
```

Here we save only 10% of both groups.
```{r}
N <- 13
res <- list()
for (i in 1:N) {
  this_sim <- muscat::simData(ref, p_dd = p_dd, nc = round(nc * 0.1), ns = 1, paired = FALSE)
  print(table(this_sim$sample_id))
  print(metadata(this_sim)$ref_sids)
  
  res[[i]] <- list(counts = Matrix(assay(this_sim, 'counts')),
                   cell_metadata = colData(this_sim),
                   sim_metadata = metadata(this_sim)[c('experiment_info', 'n_cells', 'gene_info', 'ref_sids')])
}
saveRDS(object = res, file = file.path(OUT_DIR, 'sim_data_20_vs_20.Rds'))
```

Here we create ten times as many cells for both groups.
```{r}
N <- 13
res <- list()
for (i in 1:N) {
  this_sim <- muscat::simData(ref, p_dd = p_dd, nc = nc * 10, ns = 1, paired = FALSE)
  print(table(this_sim$sample_id))
  print(metadata(this_sim)$ref_sids)
  
  res[[i]] <- list(counts = Matrix(assay(this_sim, 'counts')),
                   cell_metadata = colData(this_sim),
                   sim_metadata = metadata(this_sim)[c('experiment_info', 'n_cells', 'gene_info', 'ref_sids')])
}
saveRDS(object = res, file = file.path(OUT_DIR, 'sim_data_2k_vs_2k.Rds'))
```

An unbalanced version (200 vs 2000)

```{r}
N <- 13
res <- list()
for (i in 1:N) {
  this_sim <- muscat::simData(ref, p_dd = p_dd, nc = nc * 10, ns = 1, paired = FALSE)
  sample_tab <- table(this_sim$sample_id)
  print(sample_tab)
  print(metadata(this_sim)$ref_sids)
  
  keep <- sort(sample(x = which(this_sim$sample_id == names(sample_tab)[1]), size = round(sample_tab[1] * 0.1)))
  keep <- c(keep, which(this_sim$sample_id == names(sample_tab)[2]))
  print(table(this_sim$sample_id[keep]))
  res[[i]] <- list(counts = Matrix(assay(this_sim, 'counts'))[, keep],
                   cell_metadata = colData(this_sim)[keep, ],
                   sim_metadata = metadata(this_sim)[c('experiment_info', 'n_cells', 'gene_info', 'ref_sids')])
}
saveRDS(object = res, file = file.path(OUT_DIR, 'sim_data_200_vs_2k.Rds'))
```


# Appendix

Runtime: `r time_diff(SETUP_TIME)`

Session Info
```{r}
sessionInfo()
```
