---
title: 'Summarize results on simulated data'
author: 'Christoph Hafemeister'
date: '`r format(Sys.time(), "%B %d, %Y %H:%M:%S %Z")`'
output:
  html_document: 
    toc: yes
    toc_float: yes
    code_folding: hide
    highlight: pygments
    df_print: paged
---

# Setup

```{r}
source("R/setup.R")
source('R/utilities.R')

library('dplyr')
library('future.apply')

DATA_DIR <- file.path(config$out_root, 'simulated_data')
RES_DIR <- file.path(config$out_root, 'results', 'simulated_data')
OUT_DIR <- file.path(config$out_root, 'results', 'simulated_data_summarized')
dir.create(path = OUT_DIR, showWarnings = FALSE, recursive = TRUE)
```

# Main

For each experiment, read all results and calculate performance metrics per pipeline and replicate.

Also calculate performance stratified by gene mean:

Place genes into groups
- low expression (mean < 0.1)
- medium expression (0.1 <= mean < 1)
- high expression (1 <= mean)

```{r}
experiments <- list.files(path = DATA_DIR, full.names = FALSE) %>%
  stringr::str_remove(pattern = '\\.Rds$') %>%
  stringr::str_remove(pattern = '^sim_data_')

gene_mean_breaks <- c(-Inf, 0.1, 1, Inf)

for (experiment in experiments) {
  rds_files <- list.files(path = RES_DIR, pattern = sprintf('^%s-.*\\.Rds$', experiment), full.names = TRUE)
  
  res_list <- future_lapply(X = rds_files, FUN = function(f) {
    f_parts <- stringr::str_split(string = stringr::str_remove(basename(f), pattern = '\\.Rds$'), pattern = '-')[[1]]
    replicate <- as.numeric(f_parts[2])
    transformation <- f_parts[3]
    de_method <- f_parts[4]
    pipeline <- paste(transformation, de_method, sep = '-')
    
    res <- readRDS(f) #$res %>% mutate(file = basename(f))
    transformation_time = dplyr::filter(res$timing, step == 'transformation')$time
    method_time = dplyr::filter(res$timing, step == 'de_method')$time
    
    # add geometric mean of simulation, and pipeline label, and fix missing p-values
    # also fix inverted logFC of DP-simulated genes 
    # (bug in muscat, see https://github.com/HelenaLC/muscat/issues/125)
    de_res <- res$res %>% 
      mutate(logFC = case_when(category == 'dp' ~ -logFC, TRUE ~ logFC),
             sim_mean = expm1((log1p(sim_mean.A) + log1p(sim_mean.B)) / 2),
             expr_grp = cut(sim_mean, breaks = gene_mean_breaks, labels = c('low', 'medium', 'high')),
             pval = case_when(is.na(pval) ~ 1, TRUE ~ pval),
             FDR = p.adjust(pval, method = 'fdr'),
             pred_de = factor(FDR < 0.05, levels = c(FALSE, TRUE)),
             pred_de3 = factor(effect_direction * (FDR < 0.05), levels = c(-1, 0, 1)),
             true_de = factor(!is.na(logFC), levels = c(FALSE, TRUE)),
             true_de3 = factor(case_when(is.na(logFC) ~ 0, logFC > 0 ~ 1, logFC < 0 ~ -1), levels = c(-1, 0, 1)))
    
    performance <- summarise(de_res, perf_metrics(pred_de3, true_de3)) %>%
      mutate(time_trans = transformation_time, time_de = method_time)
    
    performance_by_expr_grp <- group_by(de_res, expr_grp) %>%
      summarise(perf_metrics(pred_de3, true_de3), .groups = 'drop')
    
    performance_by_category <- group_by(de_res, category) %>%
      summarise(perf_metrics(pred_de3, true_de3), .groups = 'drop')
    
    performance_sum <- dplyr::bind_rows(performance, performance_by_expr_grp, performance_by_category) %>%
      mutate(experiment = experiment, replicate = replicate, transformation = transformation, de_method = de_method, pipeline = pipeline)
      
    return(performance_sum)
  })
  
  res <- as_tibble(data.table::rbindlist(res_list)) %>%
    mutate(across(where(is.character), as.factor))
  
  saveRDS(res, file = file.path(OUT_DIR, sprintf('performance_%s.Rds', experiment))) 
}
```



# Appendix

Runtime: `r time_diff(SETUP_TIME)`

Session Info
```{r}
sessionInfo()
```

Future plan
```{r}
future::plan()
```

More session info
```{r}
devtools::session_info()
```
